\chapter{Related Work}
\label{sec:related_work}

This thesis sits at the meeting point of several lines of research: Bayesian experimental design, simulation-based inference, and the numerical treatment
of stiff differential equations. In addition to introducing related works, this chapter develops fundamental concepts in enough detail that the related works could be properly explained. We begin with the design objective itself
(Section~\ref{sec:rw-bed}), then with simulation-based inference
(Section~\ref{sec:rw-sbi}), which supplies the posterior estimator PACE reuses. We then trace the contrastive
estimators that policy methods train on, including the learned-proposal bounds PACE most
resembles (Section~\ref{sec:rw-policy}), the reinforcement-learning variants that share
their reward (Section~\ref{sec:rw-rl}), and the per-trajectory family that, like PACE,
uses the current posterior (Section~\ref{sec:rw-pertraj}). Two shorter sections cover the classical optimal-design criteria for ODE models and
model discrimination (Section~\ref{sec:rw-classical}) and the cost of differentiating
through stiff simulators
(Section~\ref{sec:rw-stiff}). The chapter closes by stating the gap PACE fills
(Section~\ref{sec:rw-gap}).

\section{Bayesian experimental design and the information objective}
\label{sec:rw-bed}

Bayesian experimental design chooses experiments to maximise the expected information gain
(EIG) about unknown parameters~\citep{lindley1956measure,chaloner1995bayesian,ryan2016review}.
For a design $\xi$ and a parameter $\theta$ with prior $p(\theta)$, the EIG is the mutual
information between $\theta$ and the observation $y$,
\begin{equation}
\label{eq:rw-eig}
\mathrm{EIG}(\xi)=I(\theta;y\mid\xi)
=\mathbb{E}_{p(y\mid\xi)}\!\big[H[p(\theta)]-H[p(\theta\mid y,\xi)]\big]
=\mathbb{E}_{p(\theta)p(y\mid\theta,\xi)}\!\left[\log\frac{p(y\mid\theta,\xi)}{p(y\mid\xi)}\right].
\end{equation}
The two forms provides complementing interpretation to the same objective. The middle form reads as the expected drop in entropy from the prior to the posterior, and the right form reads as an expected
log-ratio of the likelihood to the marginal. The difficulty in both is the marginal likelihood $p(y\mid\xi)=\int p(y\mid\theta,\xi)p(\theta)\ud\theta$, which is intractable
for all but the simplest models, as it requires integrating the likelihood over the entire prior distribution. 
Classical estimators use nested Monte Carlo: for each candidate design, an outer loop draws parameter–observation pairs $(\theta, y)$ from the joint, and an inner loop draws fresh parameters from the prior to approximate the marginal likelihood at each sampled $y$~\citep{ryan2016review,huan2014simulation}. \citet{rainforth2018nesting} showed that the mean squared error is $O(1/N_{\mathrm{outer}} + 1/N_{\mathrm{inner}})$, so both sample counts must grow for convergence~\citep{rainforth2018nesting}, and the total cost scales multiplicatively with the required accuracy. In the \emph{sequential} setting the design at each step depends on the data seen so far, which
turns the problem into a control problem over a growing, history-dependent design space. \citet{rainforth2024modern} provides a review of the modern landscape of these methods. The methods that dominate this sequential setting all reduce the intractable marginal to a contrastive average over sampled parameters, and they differ in where those contrastives come from and whether a policy is trained. The posterior suffers from the same intractable integral. The next section reviews the methods that approximate it.


\section{Simulation-based inference}
\label{sec:rw-sbi}

In the models we consider, the simulator can be run forward to draw samples $y\sim p(y\mid\theta,\xi)$, and the observation likelihood $p(y\mid\theta,\xi)$ can be evaluated pointwise, but the posterior $p(\theta\mid y,\xi)$ requires integrating over the full prior and has no closed form. It can be computed by MCMC, but sequential design requires a fresh posterior at every step, making per-step MCMC prohibitively expensive. Simulation-based inference (SBI) is the family of methods that approximate the posterior in models where the simulator can be run forward but the likelihood cannot be evaluated or is too expensive to evaluate repeatedly~\citep{cranmer2020frontier,deistler2025practical}. Here the likelihood is available, but SBI provides amortization: a posterior estimator trained once offline that returns cheap samples at every step without rerunning inference.

\emph{Neural posterior estimation} (NPE) parameterises the posterior $q_\phi(\theta\mid x)$ directly with a conditional density estimator and trains it by maximum likelihood on simulated pairs $(\theta,x)\sim p(\theta)p(x\mid\theta)$, which minimises the forward Kullback-Leibler divergence $\mathbb{E}_{p(x)}[\,\mathrm{KL}(p(\theta\mid x)\,\|\,q_\phi(\theta\mid x))\,]$ and so targets the true posterior at every conditioning value. \citet{papamakarios2016fast} used mixture density networks, \citet{lueckmann2017flexible} replaced them with masked autoregressive flows in a sequential scheme, and \citet{greenberg2019automatic} introduced automatic posterior transformation (NPE-C) with a correction that keeps the target exact across sequential rounds. \emph{Neural likelihood estimation} instead learns a surrogate likelihood and samples the posterior with MCMC~\citep{papamakarios2019sequential}, and \emph{neural ratio estimation} learns the likelihood-to-evidence ratio by classification~\citep{hermans2020likelihood,durkan2020contrastive,miller2022contrastive}. PACE uses NPE because it provides direct posterior samples and an evaluable posterior density, both of which the importance weighting of Chapter~\ref{sec:methodology} requires.

The density estimator inside NPE is usually a normalizing
flow~\citep{papamakarios2021normalizing}, an invertible map between a simple base distribution and the target whose exact density follows from the change-of-variables formula. We use neural spline flows~\citep{durkan2019neural} because they give fast and exact density evaluation, which the importance weights need at every step. More recent
flow-matching posterior estimators~\citep{lipman2023flow,wildberger2024flow,carrasco2026pawsterior}
train a continuous-time velocity field~\citep{chen2018neural} and scale to high dimensions, but they require an ODE solve per density evaluation, which would dominate the per-step cost of an importance-weighted acquisition score. The \texttt{sbi} toolkit packages these methods~\citep{tejerocantero2020sbi}, the \texttt{sbibm} suite benchmarks
them~\citep{lueckmann2021benchmarking}, and and architectures that handle variable observation counts~\citep{gloeckler2024all} let a single network return a posterior at every step of a growing history. Sequential SBI methods retrain the proposal across rounds to keep importance weights well behaved~\citep{deistler2022truncated}. When the proposal is not retrained, the effective sample size degrades as the posterior concentrates, at a rate governed by the Kullback-Leibler divergence between target and proposal~\citep{chatterjee2018sample}.


% Two properties of this literature shape the design of PACE. First, sequential SBI keeps its importance weights well behaved by retraining the proposal across rounds, for example with truncated proposals~\citep{deistler2022truncated}. PACE deliberately does not retrain:
% it freezes one proposal and accepts that, as the posterior concentrates, the weights degrade. The reason this is tolerable for design, and the analysis it requires, is the subject of Chapter~\ref{sec:methodology}. Second, the number of importance samples needed to control the variance of a self-normalized estimate grows like the exponential of the
% Kullback-Leibler divergence between target and proposal~\citep{chatterjee2018sample}. This
% is the formal reason a frozen proposal cannot give a calibrated EIG \emph{value} once the posterior is much tighter than the proposal, and it is why PACE is read as a design ranking rather than an absolute information estimate. Because amortized posteriors can be miscalibrated, we test ours with the standard diagnostics: trust and coverage analysis~\citep{hermans2022trust}, simulation-based calibration~\citep{talts2020validating}, distance-to-random-point coverage~\citep{lemos2023tarp}, and Pareto-smoothed importance sampling~\citep{vehtari2024pareto}.

\section{Amortized policy-based design and contrastive-EIG}
\label{sec:rw-policy}

\subsection{Contrastive EIG estimation: PCE and ACE}
\label{sec:rw-contrastive}

The variational approach to EIG estimation~\citep{foster2019variational} replaced the intractable terms in the nested Monte Carlo estimator with learned amortized approximations, yielding both lower and upper bounds on the EIG with faster convergence. A lower bound follows from learning an approximate posterior, and a consistent upper bound from learning a proposal for the inner NMC loop. \citet{foster2020unified} introduced a contrastive formulation that unifies estimation and design optimisation into a single stochastic gradient ascent. Let $\theta_0\sim p(\theta)$ generate the observation $y$, and let $\theta_{1:L}$ be contrastive parameters. \emph{Prior contrastive estimation} (PCE) draws the contrastives from the prior, $\theta_{1:L}\sim p(\theta)$, and estimates the EIG by
  \begin{equation}
  \label{eq:rw-pce}
  \hat I^{\mathrm{PCE}}_L(\xi)
  =\log\frac{p(y\mid\theta_0,\xi)}{\frac{1}{L+1}\sum_{\ell=0}^{L}
  p(y\mid\theta_\ell,\xi)} .
  \end{equation}
The denominator is a Monte Carlo estimate of the marginal $p(y\mid\xi)$, and including the data-generating $\theta_0$ in the sum makes Eq.~\eqref{eq:rw-pce} a lower bound on the EIG. \emph{Adaptive contrastive estimation} (ACE) instead draws the contrastives from a \emph{learned} proposal $q_\phi(\theta\mid y)$ and reweights each by the prior-to-proposal ratio,
  \begin{equation}
  \label{eq:rw-ace}
  \hat I^{\mathrm{ACE}}_L(\xi)
  =\log\frac{p(y\mid\theta_0,\xi)}{\frac{1}{L+1}\sum_{\ell=0}^{L}
  p(y\mid\theta_\ell,\xi)\,\dfrac{p(\theta_\ell)}{q_\phi(\theta_\ell
  \mid y)}},
  \qquad \theta_{1:L}\sim q_\phi(\theta\mid y),
  \end{equation}
so the reweighted denominator estimates the same marginal using a better-placed proposal. The weight $p(\theta_\ell)/q_\phi(\theta_\ell\mid y)$ corrects for sampling from $q_\phi$ rather than the prior. \citet{foster2020unified} establishes that ACE is a consistent lower bound, monotone in $L$, and exact for every $L$ when the proposal equals the posterior. The sequential extensions of these bounds, sPCE and sNMC, are now the standard evaluation brackets for adaptive design methods. Both PCE and ACE are static, single-experiment estimators; the sequential lift is the subject of Section~\ref{sec:rw-dad}.

Any contrastive lower bound on mutual information computed from $L{+}1$ samples is capped at $\log(L+1)$ nats~\citep{mcallester2020formal, Paninski2003EstimationOE}. For PCE this follows directly: the denominator of Eq.~\eqref{eq:rw-pce} includes the data-generating likelihood $p(y\mid\theta_0,\xi)$, so the average is at least $\frac{1}{L+1}p(y\mid\theta_0,\xi)$ and the log-ratio cannot exceed $\log(L+1)$. The same ceiling appears in the InfoNCE bound from representation learning~\citep{oord2018representation,poole2019variational}, which shares the contrastive structure. Estimators that avoid this ceiling instead have variance that grows exponentially with the true mutual information~\citep{song2020understanding}, so no feasible sample budget recovers a large information gain accurately. Chapter~\ref{sec:methodology} shows that this ceiling drives the saturation of the sPCE training signal on high-information systems.


\subsection{Sequential amortization: DAD, iDAD, and the learned-proposal bounds}
\label{sec:rw-dad}

In the sequential setting, an experiment consists of $T$ steps. At step $t$ a design $\xi_t$ is chosen based on all previous design--outcome pairs, the history $h_{t-1} = \{(\xi_i,y_i)\}_{i<t}$, and an observation $y_t \sim p(y_t \mid \theta, \xi_t)$ is collected. A design policy $\pi$ maps each history to the next design, $\xi_t = \pi(h_{t-1})$, so the full trajectory $h_T = {(\xi_t, y_t)}_{t=1}^T$ is determined jointly by $\theta$ and $\pi$. The objective is the total expected information gain $\mathcal{I}_T(\pi)$, the mutual information between $\theta$ and $h_T$ under the policy.

Deep Adaptive Design (DAD)~\citep{foster2021deep} amortizes this objective by training a policy network $\pi_\phi$ offline on simulated trajectories; at deployment the network returns a design in a single forward pass. The training signal is the sequential prior-contrastive bound (sPCE), the trajectory analogue of Eq.~\eqref{eq:rw-pce} with the per-experiment likelihood replaced by the trajectory likelihood $p(h_T\mid\theta,\pi)$. Implicit DAD (iDAD)~\citep{ivanova2021implicit} extends DAD to likelihood-free simulators by replacing the likelihood with a learned critic, and Step-DAD~\citep{hedman2025stepdad} refines a pretrained DAD policy with gradient steps at selected points during deployment, trading deployment compute for better designs.


The learned-proposal versions of the sequential bound were derived in this line of work but
never evaluated. \citet{foster2021deep} state and prove the sequential ACE bound in their
appendix but only train DAD on prior-contrastive sPCE. Written out, the sequential ACE bound
\citep[Theorem~5, Eq.~91]{foster2021deep} is
\begin{equation}
\label{eq:sace}
I_T(\pi) \ge \mathbb{E}_{p(\theta_0,h_T\mid\pi)\,q(\theta_{1:L};h_T)}
\!\left[\log\frac{p(h_T\mid\theta_0,\pi)}{\frac{1}{L+1}\sum_{\ell=0}^{L} p(h_T\mid\theta_\ell,\pi)\,\dfrac{p(\theta_\ell)}{q(\theta_\ell;h_T)}}\right],
\end{equation}
which is the trajectory analogue of Eq.~\eqref{eq:rw-ace}, with the proposal conditioning on the
full trajectory $h_T$. 
\citet{ivanova2021implicit} give the likelihood-free analogue, sLACE (their Appendix~B, Proposition~4), which replaces the trajectory likelihood $p(h_T\mid\theta_\ell,\pi)$ with a learned critic $U(h_T,\theta_\ell)$,
\begin{equation}
\label{eq:slace}
I_T(\pi) \ge \mathbb{E}_{p(\theta_0,h_T\mid\pi)\,q(\theta_{1:L};h_T)}
\!\left[\log\frac{U(h_T,\theta_0)}{\frac{1}{L+1}\sum_{\ell=0}^{L} U(h_T,\theta_\ell)\,\dfrac{p(\theta_\ell)}{q(\theta_\ell;h_T)}}\right],
\end{equation}
but the reported iDAD experiments all use its prior-proposal special case, the InfoNCE
bound. In both Eqs.~\eqref{eq:sace} and~\eqref{eq:slace} the contrastive weight is
$p(\theta_\ell)/q(\theta_\ell;h_T)$, so the denominator estimates the \emph{prior} marginal
$\int p(h_T\mid\theta,\pi)p(\theta)\ud\theta=p(h_T\mid\pi)$ of the trajectory mutual
information, and setting $q=p(\theta)$ recovers prior contrastives and the $\log(L+1)$
ceiling. The sequential learned-proposal estimator was therefore foreseen in this lineage
but left as a proven, unevaluated appendix result. PACE keeps the algebra of
Eq.~\eqref{eq:sace} but conditions the proposal on the history prefix and reweights toward
the running posterior rather than the prior, a change Chapter~\ref{sec:methodology} makes
precise and derive further.

\section{Reinforcement-learning approaches}
\label{sec:rw-rl}

A distinct line of research trains the design policy by reinforcement learning, casting
sequential design as a Markov decision process in which the state is the history, the action
is the next design, and the reward is an information measure. \citet{blau2022optimizing}
optimise the policy with deep RL against an sPCE-style information reward,
\citet{huan2016sequential} use approximate dynamic programming for the same objective, and
\citet{shen2025vsoed} introduce variational sequential optimal experimental design (vsOED),
training the policy by actor-critic RL with a reward built from a variational posterior
rather than a prior-contrastive bound. Because these methods learn the policy from sampled
rewards rather than from pathwise gradients, they \emph{avoid differentiating through the
simulator}, which is one of the two obstacles this thesis confronts. The property is not
shared by every reinforcement-learning design method: the model-aware approach of
\citet{belliardo2024modelaware} in quantum metrology backpropagates through the simulator by
automatic differentiation, and so it carries the differentiable-simulator cost that the
others avoid.

Whether the second obstacle is also avoided depends on the reward. A policy trained on the
prior-contrastive sPCE bound inherits its $\log(L+1)$ ceiling, so its signal saturates on
high-information experiments. The variational reward of vsOED is not contrastive and has no
such ceiling, so vsOED in principle avoids both obstacles at once, addressing the same two
difficulties as PACE but by training an amortized policy rather than reusing a fixed
posterior. The qualification is that its reward is tight only when the learned posterior is
accurate, which is itself hard on stiff, high-dimensional systems, and vsOED has been shown
on non-stiff benchmarks such as source location rather than the stiff ODEs studied here. We
do not compare against it, which is a limitation of our evaluation.

\section{Per-trajectory and posterior-guided design}
\label{sec:rw-pertraj}

The second dominant family does not train a policy at all. It maintains an estimate of the
current posterior and chooses each design by optimising an acquisition score against that
posterior, re-fitting as data arrive. This is the family PACE belongs to, so we are precise
about how its members work and why they have not, until now, been shown to beat amortized
policies on the hardest systems.

\citet{kleinegesse2020bayesian} estimate the EIG for implicit models with
mutual-information neural estimation and optimise the design directly.
\citet{iollo2024pasoa} and \citet{iollo2025contrastive} develop posterior-guided and
contrastive acquisition criteria, \citet{zaballa2025bridging} bridge sequential design and
SBI by optimising designs through an amortized inference model, and
\citet{pauwels2014bayesian} give an earlier instance of the same idea for systems-biology
ODEs, sequentially selecting informative conditions by active learning. The closest method
to PACE is \citet{klein2026supercharging}, who build several EIG estimators on modern SBI
density estimators and optimise them by multiple-restart gradient ascent through the
simulator, unifying inference and design. These methods share PACE's use of the current
posterior, which frees them from the $\log(L+1)$ ceiling that the prior-contrastive
objective imposes on the policy family.

Until recently this family was found to be the weaker option. The review of
\citet{rainforth2024modern} advocates amortized policies on both deployment speed and design
quality, attributing the quality gain to the policy being non-myopic and to its avoiding the
approximate per-step inference that online methods rely on; \citet{ivanova2021implicit} likewise
report that their learned policy outperforms costly non-amortized designs. The most direct
comparison is \citet{zaballa2025bridging}, who build per-trajectory design on SBI density
estimators and reach a lower information gain than policy-based methods.
\citet{klein2026supercharging} overturned part of this picture: with a multiple-restart gradient
ascent that escapes the local optima of the non-convex EIG surface, an SBI-based per-trajectory
method matches or exceeds the trained policy, which they read as evidence that the achievable
performance is higher than earlier work had found. Two costs, however, keep the family from being
the obvious choice on the systems of this thesis. First, the leading methods, Klein's included,
select designs with gradients taken through the simulator, which on stiff ODE solvers needs an
adjoint pass that is expensive, numerically delicate, or
unavailable~\citep{klein2026supercharging,fronk2024stiff,pal2023neural}. Second, several of them
retrain or update the posterior at every step, paying a per-step inference cost that amortized
policies avoid. The result is a regime that falls between the two families: on stiff,
high-information systems, policy training loses its signal to saturation while gradient-based
per-trajectory design cannot obtain reliable simulator gradients, so neither is usable. A central
empirical claim of this thesis is that a per-trajectory method which avoids both costs, by
reusing a one-time posterior estimator and scoring designs with forward simulations only, beats
amortized policies on these systems with far less computation than the gradient-based route, and
on the stiffest of them is the only feasible choice. PACE departs from the methods above by using
a posterior that is trained once and frozen, and by selecting designs without any simulator
gradient.

\section{Classical optimal design for ODE models and model discrimination}
\label{sec:rw-classical}

Optimal experimental design has a long history that predates amortization. Classical theory
selects the design that optimises a scalar functional of the Fisher information matrix,
controlling the variance of the resulting parameter estimates. \citet{smith1918standard}
gave the earliest construction, deriving designs that minimise the maximum prediction
variance for polynomial regression. The determinant criterion, which maximises $\det F$ to
shrink the volume of the confidence ellipsoid, was proposed by \citet{wald1943efficient}; it
is now known as D-optimality, and \citet{kiefer1960equivalence} proved its equivalence to
G-optimality, which minimises the worst-case prediction variance. The remaining alphabetic
criteria optimise other functionals of $F$, with A-optimality minimising the average parameter
variance and E-optimality the worst-conditioned direction. \citet{kiefer1959optimum} developed the general theory of approximate designs that
unifies these criteria, and \citet{fedorov1972theory} and \citet{atkinson1992optimum} provide
the exchange algorithms and the standard textbook treatment.

For the nonlinear models studied in this thesis, the Fisher information depends on the unknown
parameters, so a design can be optimal only locally, at an assumed parameter value.
\citet{chernoff1953locally} introduced this notion of local optimality, and
\citet{box1959design} derived D-optimal designs for nonlinear mechanistic models,
the setting closest to biochemical ODEs. This dependence on the unknown parameters is the
central difficulty: a design that is optimal at one parameter value may be poor at another,
which motivates the Bayesian and sequential treatments that average or update over parameter
uncertainty. \citet{huan2014simulation} extended optimal design to nonlinear systems through
simulation, and \citet{braniff2022nloed} provide NLoed, a Python package that constructs such
designs from the expected Fisher information matrix for systems-biology models.

When the goal is to choose between competing models rather than to estimate parameters, the
design criterion changes. \citet{hunter1965designs} proposed selecting the experiment at which
two rival models disagree most, and \citet{box1967discrimination} gave a Bayesian sequential
criterion based on the model posterior. \citet{atkinson1975discriminating2} formalised this
line as T-optimality, which maximises the lack of fit of the inferior model. In systems
biology, \citet{flassig2012optimal} designed optimal stimulus experiments that separate
competing biochemical ODE models by the overlap of their predictive distributions, and
\citet{hainy2022optimal} recast discrimination as a classification problem, selecting designs
by their misclassification error and showing that optimal designs outperform equidistant ones.

\section{Stiff dynamical systems: inference, identifiability, and design}
\label{sec:rw-stiff}

Gradient-based design optimisation and policy training by the sPCE objective both differentiate
through the ODE solver, and on stiff systems this differentiation is the dominant cost and a
source of instability. A system is stiff when its solution combines processes on widely separated
timescales~\citep{curtiss1952integration}; explicit solvers must then take very small steps to
stay stable even where the trajectory is smooth, so stiff models are integrated instead with
implicit solvers that solve a nonlinear system at every step~\citep{hairer1996solving}. Gradients
through such a solver are obtained in one of two ways. The continuous adjoint method solves an
auxiliary system backward in time and recovers the gradient at a memory cost independent of the
number of steps~\citep{chen2018neural}, whereas the discrete adjoint, or discretise-then-optimise,
differentiates the solver's own operations and returns the exact gradient of the discretised
solution at the cost of storing the forward pass.

Stiffness makes both routes fragile. The adjoint system of a stiff forward problem is itself
stiff, so it must again be solved implicitly and is numerically delicate, and the sensitivities to
fast-decaying modes shrink towards zero, a vanishing-gradient pathology that
\citet{fronk2025vanishing} show is intrinsic to stiff neural differential equations and affects
every solver damped enough to handle stiff dynamics in the first place. These difficulties
have driven a dedicated line of work on differentiable
stiff solvers: implicit single-step methods for training stiff neural ODEs~\citep{fronk2024stiff},
semi-implicit schemes~\citep{aaai2025semiimplicit}, algebraically reversible solvers that recover
the backward pass without storing the forward trajectory~\citep{mccallum2024efficient}, learned
time-reparametrisations that map a stiff system onto a non-stiff one~\citep{caldana2025neural}, and
efficient discrete-adjoint implementations~\citep{pal2023neural}.

These difficulties are individually familiar in the modelling of dynamical biochemical systems,
which the standard benchmark suites collect and standardise~\citep{hass2019benchmark}. Stiffness is
the rule rather than the exception in these models~\citep{stadter2021benchmarking}, where rate
constants routinely span several orders of magnitude, and their parameters are frequently
non-identifiable, whether structurally~\citep{villaverde2016structural} or only in
practice~\citep{raue2009identifiability}; the resulting sloppiness, in which most parameter
combinations are only loosely constrained~\citep{gutenkunst2007universally}, is the source of the
broad priors used here. For the \emph{inference} problem the cost of stiff solvers has been met
head-on: large biochemical models are routinely fitted by adjoint sensitivity analysis, which
makes the gradient cost effectively independent of the number of parameters and scales estimation
to models with hundreds of states and parameters~\citep{froehlich2017scalable}. That work,
however, estimates parameters from data already collected, not the experiments that produce the
data.

\section{Summary and the gap}
\label{sec:rw-gap}

The chapter has traced two families that dominate sequential design, and on the systems
studied here each fails for a different reason. The amortized policy family trains a network
on a single trajectory objective, but that objective is a prior-contrastive bound capped at
$\log(L+1)$, so its signal saturates on high-information experiments, and the
reinforcement-learning variants trained on the same bound saturate with it. The per-trajectory
family escapes the ceiling by scoring designs against the current posterior, but only by
differentiating through the simulator, which stiff solvers cannot do reliably, or by re-fitting
the posterior at every step. On stiff, high-information systems both obstacles are present at
once, and neither family is usable. The gap is a per-trajectory method that keeps the current
posterior but pays neither cost: it reuses a posterior trained once by simulation-based
inference and scores designs with forward simulations only. Closing this gap is the goal of
this thesis. In the next chapter, we derive PACE in detail.
