% Appendix. The main text is self-contained; this is supporting detail.

\chapter{Standard background derivations}
\label{app:background}

This appendix records the established results used in Section~\ref{sec:m-prelim}.
They are not contributions of this thesis; they are collected here so the main text
can state them briefly.

The trajectory information bound (Eq.~\eqref{eq:traj-mi}) is derived in the main text
(Section~\ref{sec:m-policy}); we record only the contrastive bounding properties and
the per-step chain rule here.

\section{Bounding properties of the contrastive estimators}
\label{app:bounds}

Recall $\mathrm{sPCE}_L$ from Eq.~\eqref{eq:spce} and $\mathrm{sNMC}_L$, which uses
the same numerator but divides the denominator sum by $L$ over the contrastives
$\theta_{1:L}$ only.

\paragraph{sPCE is a lower bound, capped at $\log(L+1)$.}
The denominator of sPCE is an average over $L+1$ likelihoods that estimates the
same marginal $p(y_t\mid\xi_t,h_{t-1})$, but it includes the data-generating
$\theta_0$, whose likelihood is systematically large for the $y_t$ it produced.
This biases the denominator up and the ratio down, so
$\mathbb{E}[\mathrm{sPCE}_L]\le I_{h_{t-1}}(\xi_t)$. Equivalently, sPCE is a
classification bound: among the $L+1$ exchangeable draws, identifying which
generated $y_t$ has success probability at most one, so
$\mathbb{E}[\mathrm{sPCE}_L]=\log(L+1)-H(\text{index}\mid\theta_{0:L},y_t)\le\log(L+1)$.

\paragraph{sNMC is an upper bound.}
The denominator $\frac1L\sum_{\ell=1}^L p(y_t\mid\theta_\ell,\xi_t)$ is an unbiased
estimate of the marginal. By Jensen's inequality the expected log of an unbiased
estimate underestimates the log of the mean, so the log-marginal is underestimated
and the ratio overestimated, giving $\mathbb{E}[\mathrm{sNMC}_L]\ge I_{h_{t-1}}(\xi_t)$.

Both biases are $O(1/L)$, so $[\mathrm{sPCE}_L,\mathrm{sNMC}_L]$ brackets the true
EIG and tightens as $L\to\infty$.

\section{Per-step chain rule}
\label{app:chain}

Let $S_t$ be the sPCE evaluated on the length-$t$ prefix and $S_0=0$. The per-step
increment $\Delta S_t=S_t-S_{t-1}$ telescopes, $\sum_{t=1}^{T}\Delta S_t=S_T$,
recovering the trajectory bound. Each $\Delta S_t$ weights the contrastive
denominator by the running posterior, so it measures the information acquired at
step $t$ within the operative posterior rather than from a fresh prior estimate.

\chapter{Linearised EIG and the escape threshold}
\label{app:lindley}

\section{The linearised information gain equals $\Lambda$}
The main text calls $\Lambda$ the Gaussian-linearised information gain; we justify the
name. For a Gaussian prior $p(\theta)=\mathcal N(\mu_0,\Sigma_0)$ and a linear-Gaussian
observation $y=\mu_0+J(\theta-\theta_0)+\varepsilon$ with $\varepsilon\sim\mathcal N(0,R)$,
the posterior is Gaussian with covariance $\Sigma_{\mathrm{post}}=(\Sigma_0^{-1}+F)^{-1}$,
$F=J^\top R^{-1}J$, which does not depend on the realised $y$. The information gain is the
drop in differential entropy,
\begin{align}
\mathrm{EIG}(\xi)
&= H[p(\theta)]-\mathbb{E}_y\,H[p(\theta\mid y,\xi)]
= \tfrac12\log\!\big((2\pi e)^d\det\Sigma_0\big)-\tfrac12\log\!\big((2\pi e)^d\det\Sigma_{\mathrm{post}}\big)\nonumber\\
&= \tfrac12\log\frac{\det\Sigma_0}{\det\Sigma_{\mathrm{post}}}
= \tfrac12\log\det\!\big(\Sigma_0(\Sigma_0^{-1}+F)\big)
= \tfrac12\log\det(I+\Sigma_0 F)
= \tfrac12\sum_i\log(1+\lambda_i)=\Lambda. \label{eq:eig-lambda}
\end{align}
So $\Lambda$ is exactly the EIG of the linearised experiment (equivalently the Bayesian
D-optimal value). For nonlinear models the same expression is the leading term of the
entropy-based EIG under a Gauss--Newton (Laplace) approximation, with $\Sigma_{\mathrm{post}}$
evaluated at the posterior mode. Combined with the escape result below, this gives the
headline relation $L_{\mathrm{escape}}\sim e^{\Lambda}=e^{\mathrm{EIG}}$: the contrastive
budget needed to escape saturation grows as the exponential of the experiment's
information.

\paragraph{Interpretation.} Each $\lambda_i$ is a per-axis signal-to-noise ratio. It is
an eigenvalue of the prior-scaled Fisher information $\Sigma_0^{1/2}F\Sigma_0^{1/2}$,
comparing how much the experiment informs direction $i$ (through $F$) against the prior
uncertainty there (through $\Sigma_0$). A well-identified direction has $\lambda_i\gg1$
and contributes $\tfrac12\log(1+\lambda_i)\approx\tfrac12\log\lambda_i$ nats; a
barely-informed direction has $\lambda_i\ll1$ and contributes almost nothing. $\Lambda$
is the sum of these per-axis contributions, and $e^{\Lambda}=\prod_i(1+\lambda_i)^{1/2}$
is the factor by which the posterior volume shrinks below the prior, which is exactly
the contrastive budget needed to escape saturation.

\section{The expected importance weight and the escape threshold}
We derive in full the scaling $\mathbb{E}[r_\ell]\sim e^{-\Lambda}$ stated in
Eq.~\eqref{eq:escape-scale}, via the marginal-likelihood route. Take a Gaussian prior
$p(\theta)=\mathcal N(\mu_0,\Sigma_0)$ and observation $y=\mu(\theta,\xi)+\varepsilon$
with $\varepsilon\sim\mathcal N(0,R)$. The Fisher information is $F=J^\top R^{-1}J$
with $J=\partial\mu/\partial\theta|_{\theta_0}$, and the posterior precision is
$\Sigma_{\mathrm{post}}^{-1}=\Sigma_0^{-1}+F$.

\paragraph{Step 1: Laplace approximation to the marginal likelihood.}
Expand the log-likelihood to second order about $\theta_0$,
\[
\log p(y\mid\theta,\xi) \approx \log p(y\mid\theta_0,\xi) + (\theta-\theta_0)^\top s - \tfrac12(\theta-\theta_0)^\top F(\theta-\theta_0),
\]
where $s=\nabla_\theta\log p(y\mid\theta,\xi)|_{\theta_0}$ is the score and the
curvature is the observed information $H=-\nabla_\theta^2\log p(y\mid\theta,\xi)|_{\theta_0}$;
under Gaussian noise $H=J^\top R^{-1}J=F$ exactly, and for general likelihoods
$\mathbb{E}_y[H]=F$, so we use $F$ throughout. Adding the Gaussian log-prior and
completing the square,
\[
\log\bigl[p(y\mid\theta,\xi)\,p(\theta)\bigr] \approx \log\bigl[p(y\mid\theta_0,\xi)\,p(\theta_0)\bigr] + \tfrac12 g^\top\Sigma_{\mathrm{post}}\,g - \tfrac12(\theta-\hat\theta)^\top\Sigma_{\mathrm{post}}^{-1}(\theta-\hat\theta),
\]
where $g=s-\Sigma_0^{-1}(\theta_0-\mu_0)$ is the log-posterior gradient at $\theta_0$
and $\hat\theta=\theta_0+\Sigma_{\mathrm{post}}\,g$ is the posterior mode. The
unnormalised posterior is thus approximately Gaussian with covariance
$\Sigma_{\mathrm{post}}$; integrating the Gaussian kernel
($\int = (2\pi)^{d/2}\det(\Sigma_{\mathrm{post}})^{1/2}$),
\[
p(y\mid\xi) \approx p(y\mid\theta_0,\xi)\,p(\theta_0)\,e^{\frac12 g^\top\Sigma_{\mathrm{post}}\,g}\,(2\pi)^{d/2}\det(\Sigma_{\mathrm{post}})^{1/2}.
\]

\paragraph{Step 2: Expected importance weight.}
From Eq.~\eqref{eq:Er}, $\mathbb{E}[r_\ell]=\int p(y\mid\theta,\xi)\,p(\theta)\,\ud\theta\,/\,p(y\mid\theta_0,\xi)$.
Substituting the Laplace marginal and the Gaussian prior density
$p(\theta_0)=(2\pi)^{-d/2}\det(\Sigma_0)^{-1/2}\exp(-\tfrac12(\theta_0-\mu_0)^\top\Sigma_0^{-1}(\theta_0-\mu_0))$,
\[
\mathbb{E}[r_\ell] \approx \underbrace{\frac{\det(\Sigma_{\mathrm{post}})^{1/2}}{\det(\Sigma_0)^{1/2}}}_{\text{(1) volume ratio}} \cdot \underbrace{\exp\!\Bigl(-\tfrac12(\theta_0-\mu_0)^\top\Sigma_0^{-1}(\theta_0-\mu_0)\Bigr)}_{\text{(2) prior centering}} \cdot \underbrace{e^{\frac12 g^\top\Sigma_{\mathrm{post}}\,g}}_{\text{(3) mode shift}}.
\]
Factor (1) is the exponentially small volume ratio that drives saturation. Factor (2)
involves a Mahalanobis distance that is $\chi^2_d$ with mean $d$, so it is
$e^{-\Theta(d)}$ (at the mean exponent, $e^{-d/2}$; in exact expectation
$\mathbb{E}[e^{-\frac12\chi^2_d}]=2^{-d/2}$). Factor (3) is $\ge1$ with exponent
$O(d)$. Factors (2) and (3) push in opposite directions, contribute $O(d)$ to the
exponent, and do not affect the exponential scaling in $\Lambda$. Retaining the
leading volume ratio,
\[
\mathbb{E}[r_\ell]\sim\frac{\det(\Sigma_{\mathrm{post}})^{1/2}}{\det(\Sigma_0)^{1/2}}=\frac{1}{\det(I+\Sigma_0 F)^{1/2}}=\prod_i(1+\lambda_i)^{-1/2}=e^{-\Lambda}.
\]

\paragraph{Step 3: Eigenvalue identity.}
$\Sigma_0 F$ is similar to the symmetric positive semidefinite matrix
$M:=\Sigma_0^{1/2}F\Sigma_0^{1/2}$ through $P=\Sigma_0^{1/2}$,
\[
P\,M\,P^{-1}=\Sigma_0^{1/2}\bigl(\Sigma_0^{1/2}F\Sigma_0^{1/2}\bigr)\Sigma_0^{-1/2}=\Sigma_0 F .
\]
Similar matrices share eigenvalues, and $M$ is symmetric PSD, since for any $v$,
$v^\top M v=(\Sigma_0^{1/2}v)^\top F(\Sigma_0^{1/2}v)\ge0$; its eigenvalues
$\lambda_i\ge0$ are therefore real. Hence
\[
\det(I+\Sigma_0 F)=\det(I+M)=\prod_i(1+\lambda_i),
\qquad \Lambda=\tfrac12\sum_i\log(1+\lambda_i).
\]

\paragraph{Step 4: Escape threshold.}
The expected contrastive sum is $\mathbb{E}[S]=L\,e^{-\Lambda}$. By Markov's
inequality $\Pr(S\ge1)\le\mathbb{E}[S]=L\,e^{-\Lambda}$, so saturation persists while
$\mathbb{E}[S]\ll1$, i.e.\ $L\ll e^{\Lambda}$, and escape requires
$L\gtrsim e^{\Lambda}$. $\square$

\paragraph{Correction factors.}
The combined effect of factors (2) and (3) shifts $L_{\mathrm{escape},50}$ by a
multiplicative factor $\sim e^{O(d)}$, large in absolute terms for $d=12$ (ASM1) but
subexponential relative to $e^{\Lambda}$ when $\Lambda\gg d$. Table~\ref{tab:escape-full}
reports the full twelve-system measurements; the structural mechanisms that push the
\emph{empirical} threshold beyond this Laplace prediction are derived in the main text
(Section~\ref{sec:m-mechanisms}).

\begin{table}[h]
\centering\small
\caption[Full saturation thresholds]{Empirical escape thresholds at random designs,
sorted ascending, with the Laplace reference $e^\Lambda$ where available. Stiff and
Hill-kinetic systems exceed $e^\Lambda$ by one to three orders of magnitude.}
\label{tab:escape-full}
\begin{tabular}{lccrrl}
\toprule
System & $d_\theta$ & $d_\xi$ & $\Lambda$ & $e^{\Lambda}$ & $L_{\mathrm{escape},50}$ \\
\midrule
PK 8-comp & 20 & 1 & $0.05$ & $1.1$ & ${\le}10$ \\
PK 1-comp & 3 & 1 & $1.02$ & $2.8$ & ${\sim}10$ \\
PI3K      & 6 & 2 & $1.49$ & $4.4$ & ${\le}10$ \\
Monod     & 2 & 1 & $4.30$ & $74$ & ${\sim}30$ \\
CES       & 4 & 6 & -- & -- & ${\sim}10^4$ (static) \\
Goodwin   & 5 & 1 & $4.40$ & $81$ & ${\sim}10^4$ \\
ASM1      & 12 & 2 & -- & -- & ${\sim}10^4$ (stiff) \\
SEIR R=3  & 9 & 9 & $8.64$ & $5{,}648$ & ${>}10^6$ \\
SEIR R=8  & 14 & 24 & $21.26$ & $1.7{\times}10^9$ & ${>}10^6$ \\
\bottomrule
\end{tabular}
\end{table}

\chapter{PACE: consistency, decomposition, and the Fisher limit}
\label{app:pace}

\section{Consistency}
Theorem~\ref{thm:consistency} is the standard two-stage self-normalized importance
sampling result. Support inclusion makes the weighted averages in the numerator and
denominator of Eq.~\eqref{eq:pace} converge almost surely to the corresponding
posterior expectations; finite chi-square divergence gives the $O(1/J)$ bias rate
through the SNIS central limit theorem. With a perfect proposal the weights are the
constant of Eq.~\eqref{eq:widentity} and the estimator is exact in the weights.

\section{Finite variance and the likelihood-factor cancellation}
\label{app:chi2}
The $O(1/J)$ rate in Theorem~\ref{thm:consistency} requires the chi-square divergence
$\chi^2(p(\cdot\mid h)\,\|\,q_\phi(\cdot\mid h))=\mathbb{E}_{q_\phi}\!\big[(p(\theta\mid h)/q_\phi(\theta\mid h))^2\big]-1$
to be finite, which holds for any NPE that covers the posterior. It is worth seeing why
PACE's weight has this property when the naive alternative does not. Estimating the
per-step gain by drawing contrastives from $q_\phi$ but \emph{targeting the prior
marginal}, as a prior-contrastive denominator does, would weight each draw by
$p(\theta_\ell)/q_\phi(\theta_\ell\mid h)$, whose divergence is
\begin{equation}
\label{eq:chi2-prior}
\chi^2\!\big(p(\theta)\,\big\|\,q_\phi(\cdot\mid h)\big)=\int\frac{p(\theta)^2}{q_\phi(\theta\mid h)}\,\ud\theta-1 .
\end{equation}
Since $q_\phi(\cdot\mid h)$ approximates the concentrated posterior, it has negligible
mass in the prior-dominated tails where $p(\theta)$ does not vanish, so the integrand
$p(\theta)^2/q_\phi(\theta\mid h)$ diverges and this $\chi^2$ is unbounded. PACE's weight
is instead
\[
w_j=\frac{p(\theta_j)\,p(h\mid\theta_j)}{q_\phi(\theta_j\mid h)}\;\propto\;\frac{p(\theta_j\mid h)}{q_\phi(\theta_j\mid h)} ,
\]
where the likelihood factor $p(h\mid\theta_j)$ is small exactly in those prior tails and
cancels the divergence, leaving the posterior-to-proposal ratio whose $\chi^2$ is finite
for a covering NPE. This cancellation is what makes the cost transfer of
Section~\ref{sec:m-costtransfer} work: the same likelihood factor that retargets the
estimator from the prior to the posterior is what keeps its variance finite.

\section{Decomposition}
Substituting $\tilde w_k=w_k/W$ with $W=\sum_k w_k$ into Eq.~\eqref{eq:pace} and
factoring $p(y_{t,j}\mid\theta_j,\xi_t)$ out of the inner sum using
$r_{k\leftarrow j}=p(y_{t,j}\mid\theta_k,\xi_t)/p(y_{t,j}\mid\theta_j,\xi_t)$ gives,
for the $j$-th term, $\log W-\log w_j-\log[1+\Psi_j(\xi_t)]$ with
$\Psi_j(\xi_t)=\sum_{k\ne j}(w_k/w_j)r_{k\leftarrow j}$. Taking the weighted sum and
using $\sum_j\tilde w_j\log w_j=-H(\tilde w)+\log W$ yields Eq.~\eqref{eq:decomp},
$\hat I^{\mathrm{PACE}}=H(\tilde w)-\bar\Phi(\xi_t)$. The collapsed-weight limit of
Eq.~\eqref{eq:lr-limit} follows as in Section~\ref{sec:m-decomp}.

\section{The Fisher limit}
We expand the collapsed-weight criterion of Section~\ref{sec:m-collapse}. When a single
competitor $\theta_{k^\ast}=\arg\max_{k\ne\mathrm{dom}}w_k$ dominates the remaining
non-dominant particles, the design-dependent term reduces to one summand,
\begin{equation}
\label{eq:pairwise}
\bar\Phi(\xi_t)\approx\frac{w_{k^\ast}}{w_{\mathrm{dom}}}\,r_{k^\ast\leftarrow\mathrm{dom}}(\xi_t)
=\frac{w_{k^\ast}}{w_{\mathrm{dom}}}\,\frac{p(y_{t,\mathrm{dom}}\mid\theta_{k^\ast},\xi_t)}{p(y_{t,\mathrm{dom}}\mid\theta_{\mathrm{dom}},\xi_t)} .
\end{equation}
The prefactor $w_{k^\ast}/w_{\mathrm{dom}}$ is constant in $\xi_t$, so minimising
$\bar\Phi$ is maximising the pairwise log likelihood-ratio
$\log[p(y_{t,\mathrm{dom}}\mid\theta_{\mathrm{dom}},\xi_t)/p(y_{t,\mathrm{dom}}\mid\theta_{k^\ast},\xi_t)]$.
Its expectation over $y_{t,\mathrm{dom}}\sim p(y_t\mid\theta_{\mathrm{dom}},\xi_t)$ is the
Kullback-Leibler divergence
\[
\mathbb{E}_{y_{t,\mathrm{dom}}}\!\left[\log\frac{p(y_{t,\mathrm{dom}}\mid\theta_{\mathrm{dom}},\xi_t)}{p(y_{t,\mathrm{dom}}\mid\theta_{k^\ast},\xi_t)}\right]
= D_{\mathrm{KL}}\!\big(p(y_t\mid\theta_{\mathrm{dom}},\xi_t)\,\|\,p(y_t\mid\theta_{k^\ast},\xi_t)\big) .
\]
For Gaussian likelihoods with a linearised forward map
$\mu(\theta,\xi_t)\approx\mu(\theta_{\mathrm{dom}},\xi_t)+J_{\xi_t}(\theta-\theta_{\mathrm{dom}})$,
this reduces to
$\tfrac12(\theta_{\mathrm{dom}}-\theta_{k^\ast})^\top F(\xi_t)(\theta_{\mathrm{dom}}-\theta_{k^\ast})$
with $F(\xi_t)=J_{\xi_t}^\top R^{-1}J_{\xi_t}$, and treating $\theta_{k^\ast}$ as a
posterior draw with covariance $\Sigma_{\mathrm{post}}$ gives the expected criterion
$\tfrac12\,\mathrm{tr}(F(\xi_t)\Sigma_{\mathrm{post}})$, a trace approximation to the
Gaussian-linearised (D-optimal) EIG.

This KL connection holds only for the pairwise case. The general weighted-sum criterion
of Eq.~\eqref{eq:lr-limit} is a sum of likelihood \emph{ratios}, not log-ratios, and
their expectations are constant in the design,
$\mathbb{E}_{y_{t,\mathrm{dom}}}[r_{k\leftarrow\mathrm{dom}}]=\int p(y\mid\theta_k,\xi_t)\,\ud y=1$;
there it is the realised $y_{t,\mathrm{dom}}$, not its expectation, that carries the
design signal.

\section{The dominant particle, and why standard IS fixes do not help}
\label{app:domparticle}
Under a collapsed weight set the dominant particle $\theta_{\mathrm{dom}}$ is the
posterior draw that the NPE most \emph{undercovers}: its self-normalised weight
$\tilde w_j\propto p(\theta_j\mid h)/q_\phi(\theta_j\mid h)$ is largest where the proposal
places too little mass relative to the true posterior, typically a posterior-tail
sample. This is why design selection keeps a signal even as $\mathrm{ESS}\to1$: the
criterion of Eq.~\eqref{eq:lr-limit} discriminates this tail representative from the
bulk rather than collapsing onto the posterior mean.

Two natural alternatives are worse. Dropping the importance weights and scoring with the
raw NPE particles fails because, once the posterior is tight, all $J$ particles cluster
near $\theta_{\mathrm{dom}}$, every ratio $r_{k\leftarrow j}\approx1$, and the score is
flat across designs; the reweighting is what keeps a discriminative contrastive in the
bank. Standard remedies for heavy-tailed importance weights also do not apply here.
Pareto-smoothed importance sampling~\citep{vehtari2024pareto} assumes a finite-variance
tail (Pareto $\hat k<0.7$), whereas here $\hat k$ is large by construction; truncating
the weights trades the (cancelling) variance for a bias that dominates at
$\mathrm{ESS}/J\sim10^{-3}$; and defensive mixtures that fold the prior back into the
proposal reintroduce exactly the prior-contrastive cost $L_{\mathrm{escape}}$ that PACE
removes. PACE instead accepts the collapse and relies on common-mode cancellation, which
requires only that the weights are shared across candidate designs.

\chapter{Supplementary tables}
\label{app:supp}

Table~\ref{tab:ess-full} reports the deployment-time effective sample size across
steps. It declines on every system but does not track the performance gap: see
Section~\ref{sec:r-perstep}. For the per-step increments $\Delta S_t$, the
cumulative $\sum_t\Delta S_t$ for PACE is $+13.06$ on CES (DAD $+12.05$, random
$+9.05$) and $+14.83$ on SEIR R=3 (DAD $+7.46$, random $+14.15$), and exceeds random
on every system in the suite.

\begin{table}[h]
\centering\footnotesize
\caption[Effective sample size trajectories]{Median per-step effective sample size from
production rollouts ($J=1000$; reported at $J=1000$ for ASM1 for comparability). The
decline is monotone from the first step and universal; all eight systems reach
$\mathrm{ESS}\le3$ by the final step, yet the decline does not track the performance
gap (Section~\ref{sec:r-perstep}). Only PK 8-comp (omitted, $\mathrm{ESS}/J\approx12$
to $23\%$ throughout) retains a high effective sample size.}
\label{tab:ess-full}
\begin{tabular}{l c >{\raggedright\arraybackslash}p{8.2cm} l}
\toprule
System & $T$ & ESS per step (median) & $\Delta_{\mathrm{sPCE}}$ \\
\midrule
Monod & 14 & 841, 612, 330, 125, 68, 39, 36, 18, 17, 10, 5, 3, 2, 1 & $-0.66$ vs DAD \\
CES & 10 & 9, 2, 2, 2, 3, 2, 2, 3, 2, 2 & $+0.93$ vs DAD \\
Src.\ loc.\ $p{=}2$ & 10 & 40, 29, 15, 12, 6, 3, 2, 2, 2, 1 & $+1.52$ vs DAD \\
Src.\ loc.\ $p{=}5$ & 30 & 93, 75, 39, 37, 32, 29, 21, 18, 13, 11, 8, 7, 6, 5, 6, 5, 4, 4, 4, 3, 3, 3, 3, 3, 3, 2, 2, 2, 2, 2 & $+1.07$ vs DAD \\
PI3K & 8 & 561, 285, 87, 17, 8, 3, 3, 2 & $-0.14$ vs DAD \\
SEIR R=3 & 10 & 20, 8, 3, 2, 2, 1, 1, 2, 1, 1 & $+7.31$ vs DAD \\
ASM1 & 5 & 77, 16, 6, 3, 3 & $+0.89$ vs rand. \\
Goodwin & 10 & 169, 81, 37, 23, 12, 8, 6, 5, 6, 2 & $-0.48$ vs rand. \\
\bottomrule
\end{tabular}
\end{table}

\input{sections/perstep_decomposition}

\chapter{Diagnostic protocols}
\label{app:diagnostics}

\paragraph{Bootstrap pick-stability.}
On ASM1, for several steps of a ten-step rollout, draw $J=500$ particles,
bootstrap-resample $B=200$ times, recompute the PACE scores for $50$ candidates, and
record the top-1 agreement. The median is $45\%$, about $23\times$ the chance rate
of $1/50$. Residual variation comes from resampling changing which particle
dominates.

\paragraph{ESS-regime interpolation.}
Classifying each (system, step) cell by $\mathrm{ESS}/J$ as EIG-estimation
($\ge10\%$), transition ($1$--$10\%$), or dominant-particle ($<1\%$),
PACE spans a more than hundredfold range across systems without any user intervention,
and individual systems traverse the regimes within a single rollout as the posterior
tightens (Table~\ref{tab:ess-regime}). PK 8-comp, with tight priors, is the only system
that stays in the EIG-estimation regime throughout; PI3K traverses all three;
source-location and ASM1 reach the dominant-particle regime by their final steps yet
still deliver positive design value. This is the smooth interpolation of
Section~\ref{sec:m-cancellation} made empirical.

\begin{table}[h]
\centering\small
\caption[ESS regimes]{Range of $\mathrm{ESS}/J$ across steps and the regime traversed,
classified as EIG-estimation ($\ge10\%$), transition ($1$--$10\%$), or
dominant-particle ($<1\%$). The regime is set by system informativeness (the
prior-to-posterior volume ratio), not parameter dimension.}
\label{tab:ess-regime}
\begin{tabular}{lccl}
\toprule
System & $d_\theta$ & $\mathrm{ESS}/J$ range & Regime traversed \\
\midrule
PK 8-comp & 20 & 12--23\% & EIG estimation throughout \\
PI3K & 6 & 0.2--56\% & EIG $\to$ transition $\to$ dominant-particle \\
Src.\ loc.\ $p{=}2$ & 4 & 0.1--4\% & transition $\to$ dominant-particle \\
ASM1 & 12 & 0.3--8\% & transition $\to$ dominant-particle \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Raw versus resampled coverage.}
Computed in Section~\ref{sec:r-diag}, Table~\ref{tab:coverage}: raw NPE coverage of
the truth stays in $75$ to $90\%$ while resampled coverage collapses at late steps.

\paragraph{Per-step calibration.}
Table~\ref{tab:sbc} reports simulation-based calibration KS distances on raw NPE
samples and on resampled samples. Raw KS is flat or improving across steps; the
resampled KS degrades, confirming the resampling artifact rather than a proposal
failure. Simulation-based calibration tests average-case calibration; the raw
coverage measured at deployment histories is the complementary conditional check.

\begin{table}[h]
\centering\small
\caption[Per-step calibration]{Simulation-based calibration KS distance (smaller is
better) on raw NPE samples (tests $q_\phi$ directly) versus importance-resampled
samples. Upper block: stiff-ODE systems with severe ESS collapse ($n{=}20$ synthetic
trajectories, $J{=}1000$). Lower block: unsaturated systems from controlled diagnostics
(IS KS not computed). Raw KS stays low and stable across steps; IS KS degrades as the
effective sample size collapses. $^\ddagger$The source-location $p{=}8$ diagnostic uses
a $15$-step horizon; the production system uses $T{=}30$.}
\label{tab:sbc}
\begin{tabular}{lcccc}
\toprule
System & step & ESS & raw KS & IS KS \\
\midrule
CES ($d_\theta{=}4$, $T{=}10$)
  & 1 & $65.8$ & $0.17$ & $0.20$ \\
  & 5 & $6.7$ & $0.14$ & $0.33$ \\
  & 10 & $10.7$ & $0.13$ & $0.39$ \\
ASM1 ($d_\theta{=}12$, $T{=}5$)
  & 1 & $97.6$ & $0.18$ & $0.18$ \\
  & 3 & $14.1$ & $0.17$ & $0.20$ \\
  & 5 & $8.4$ & $0.17$ & $0.24$ \\
\midrule
PI3K ($d_\theta{=}6$, $T{=}8$)
  & 1 & -- & $0.14$ & -- \\
  & 8 & -- & $0.13$ & -- \\
PK 8-comp ($d_\theta{=}20$, $T{=}5$)
  & 1 & -- & $0.16$ & -- \\
  & 5 & -- & $0.15$ & -- \\
Src.\ loc.\ $p{=}8$ ($d_\theta{=}16$)$^\ddagger$
  & 1 & -- & $0.17$ & -- \\
  & 15 & -- & $0.16$ & -- \\
\bottomrule
\end{tabular}
\end{table}

\chapter{Wall-clock cost}
\label{app:walltime}

Table~\ref{tab:walltime} compares the one-time training cost and the per-step
deployment cost. PACE trains its posterior estimator once in $3.6$ to $12.8$ minutes
and then selects each design in under four seconds on eleven of twelve systems (ASM1,
a stiff $13$-state ODE, is the exception at $40.5$\,s). Policy training ranges from
minutes to sixteen hours and is infeasible on the stiffest systems.

\paragraph{Hardware.} All Snellius runs use a single NVIDIA A100-SXM4-40GB on the
\texttt{gpu\_a100} partition; this covers every PACE per-step measurement, all DAD
training, and all NPE training except ASM1 and source-location, whose NPEs were trained
on a local desktop GPU (marked $\dagger$; an A100 would be faster, which only
strengthens the comparison). Drawing $J{=}1000$ particles from a trained NPE takes
about $11$\,ms, negligible against the ODE-solve cost.

\begin{table}[h]
\centering\small
\caption[Wall-clock cost]{Per-step design time, one-time NPE training, and one-time
policy training, with the experiment timescale for context. A dagger marks a local-GPU
measurement; an asterisk marks unstable DAD training (CES diverged to NaN before
convergence; the reported time is a partial run).}
\label{tab:walltime}
\begin{tabular}{lrrrll}
\toprule
System & $T$ & PACE/step & NPE train & DAD train & Expt.\ timescale \\
\midrule
CES & 10 & $0.09$\,s & $3.7$\,min & $9$\,min$^*$ & minutes \\
ASM1 & 5 & $40.5$\,s & $12.8$\,min$^\dagger$ & infeasible & hours \\
SEIR R=3 & 10 & ${\sim}1$\,s & $3.6$\,min & $9$\,min (useless) & days--weeks \\
Monod & 14 & $0.22$\,s & $7.2$\,min & $16$\,hr & hours \\
Goodwin & 10 & $0.16$\,s & $5.1$\,min & failed & -- \\
Src.\ loc.\ $p{=}2$ & 10 & $0.13$\,s & $3$--$12$\,min$^\dagger$ & $20$\,min & -- \\
Src.\ loc.\ $p{=}5$ & 30 & $0.15$\,s & $3$--$12$\,min$^\dagger$ & $1.5$\,hr & -- \\
Src.\ loc.\ $p{=}8$ & 30 & $0.16$\,s & $3$--$12$\,min$^\dagger$ & $1.5$\,hr & -- \\
PK 1-comp & 5 & $1.21$\,s & $8.5$--$9$\,min & $7$\,min & hours \\
PK 8-comp & 5 & $0.34$\,s & $9.6$--$11$\,min & $31$\,min & hours \\
PI3K & 8 & $3.34$\,s & $8$--$9$\,min & $3.2$\,hr & hours \\
\bottomrule
\end{tabular}
\end{table}

\chapter{System specifications}
\label{app:systems}

Every system used in the thesis is defined here, one section each, in a common format:
the forward model and its constants, the prior, and the observation and design spaces.
Forward models, parameters, priors, and noise are taken from the implementation;
per-method experiment configurations (horizon $T$, particle count $J$, candidates
$N_{\mathrm{cand}}$, CEM rounds $C$, evaluation budget $L_{\mathrm{eval}}$) are collected
in Table~\ref{tab:config}. Dynamic systems are integrated with an explicit (Tsit5)
solver, except where stiffness is noted. Throughout, $\mathcal N$ denotes a normal
distribution and $\mathrm{LogNormal}(\log m,\sigma^2)$ a variable whose natural log is
$\mathcal N(\log m,\sigma^2)$.

\section{CES (constant elasticity of substitution)}
The CES task is a static economic-choice model: a respondent compares two baskets of
goods and reports a censored preference score. It is a standard adaptive-design benchmark
across the DAD family~\citep{foster2021deep,blau2022optimizing,hedman2025stepdad}, and it
has no dynamics, since each of the $T$ steps is an independent query at a new design.

\textbf{Forward model.} A design $\xi=(d_1,d_2)$ presents two baskets, each a vector of
three nonnegative quantities $d_1,d_2\in[0.01,100]^3$, so $d_\xi=6$. Given parameters
$\theta=(\rho,\alpha_1,\alpha_2,\log u)$, the respondent's CES utility of a basket
$d\in\mathbb{R}^3_{\ge 0}$ is
\[
U(d)=\Big(\textstyle\sum_{i=1}^3\alpha_i\,d_i^{\rho}\Big)^{1/\rho},
\]
where $\rho\in(0,1]$ is the elasticity of substitution and
$\alpha=(\alpha_1,\alpha_2,\alpha_3)$ are nonnegative preference weights summing to one.
The latent preference for basket~$1$ over basket~$2$ is the scaled, clipped utility
difference $\eta=\mathrm{clip}\big(u\,[U(d_1)-U(d_2)],-20,20\big)$ with sensitivity
$u=e^{\log u}>0$, and the recorded response is a censored sigmoid-normal draw
\[
y=\mathrm{clip}\big(\mathrm{sigmoid}(\eta+s\,\epsilon),\,\varepsilon,\,1-\varepsilon\big),
\qquad \epsilon\sim\mathcal N(0,1),\qquad
s=u\,\sigma_0\big(1+\|d_1-d_2\|_2\big),
\]
with base noise $\sigma_0=0.005$ and censoring level $\varepsilon=2^{-22}$. The
observation is the scalar $y\in[\varepsilon,1-\varepsilon]$, so $d_y=1$.

\textbf{Prior.} The elasticity is $\rho\sim\mathrm{Uniform}(0.01,1)$; the preference
weights are $\alpha\sim\mathrm{Dirichlet}(1,1,1)$ (uniform on the two-simplex), of which
only the free pair $(\alpha_1,\alpha_2)$ is stored, with $\alpha_3=1-\alpha_1-\alpha_2$;
and the log-sensitivity is $\log u\sim\mathcal N(1,3^2)$. In total $d_\theta=4$. The
probability mass that the sigmoid-normal places near the censoring boundaries is the
heavy-tailed likelihood feature that destabilises gradient-based policy training here.

\section{Source location}
A static geolocation task: two hidden point sources emit a signal, and each design places
one sensor whose reading is the noisy, log-compressed superposition of the source
intensities. It is the standard scalable BED
benchmark~\citep{foster2021deep,klein2026supercharging}, with the per-step design
dimension $p$ controlling difficulty.

\textbf{Forward model.} The two sources are $\theta=(\theta_1,\theta_2)$ with
$\theta_k\in\mathbb{R}^p$, and a sensor at design $\xi\in[-3,3]^p$ ($d_\xi=p$) reads the
field at $M=3$ resolution scales. At scale $m$ the noiseless reading is the log-intensity
\[
G_m(\xi,\theta)=\log\!\Big(b+\sum_{k=1}^{2}\frac{1}{\sigma_m^2+\|\xi-\theta_k\|_2^2}\Big),
\qquad m=1,2,3,
\]
with baseline $b=0.1$ and bandwidths $\sigma_1^2=0.001$ (sharp), $\sigma_2^2=0.1$
(medium), $\sigma_3^2=1.0$ (broad): the sharp kernel localises a nearby source while the
broad kernel responds to all sources. Each channel has independent additive noise,
\[
y_m=G_m(\xi,\theta)+\varepsilon_m,\qquad \varepsilon_m\sim\mathcal N(0,0.5^2),
\]
so the per-step observation has $d_y=M=3$ channels.

\textbf{Prior.} Each source is standard normal, $\theta_k\sim\mathcal N(0,I_p)$, and the
two sources are reordered by increasing distance from the origin
($\|\theta_1\|_2\le\|\theta_2\|_2$) to break the label-switching symmetry, following
\citet{klein2026supercharging}. Hence $d_\theta=2p$.

\textbf{Variants.} We report three sizes, $p\in\{2,5,8\}$, giving
$(d_\theta,d_\xi)=(4,2),(10,5),(16,8)$ with horizons $T=10,30,30$ respectively
(Table~\ref{tab:config}).

\section{Pharmacokinetics: one compartment}
The iDAD one-compartment pharmacokinetic model~\citep{ivanova2021implicit}: a single dose
is absorbed and eliminated, and each design chooses a blood-sampling time at which the
plasma concentration is measured. The concentration has a closed form, so no ODE solve is
needed.

\textbf{Forward model.} With absorption rate $k_a$, elimination rate $k_e$, and volume of
distribution $V$, so that $\theta=(k_a,k_e,V)$ and $d_\theta=3$, the plasma concentration
a time $\xi\in[0,24]$\,h after a fixed dose $D=400$ is the closed form
\[
C(\xi,\theta)=\frac{D}{V}\,\frac{k_a}{k_a-k_e}\big(e^{-k_e\xi}-e^{-k_a\xi}\big).
\]
The measurement is heteroscedastic Gaussian, with a proportional and an additive part in
the standard deviation,
\[
y\sim\mathcal N\!\big(C,\,s^2\big),\qquad s=\sqrt{(0.1\,C)^2+0.1},
\]
that is, ten percent relative noise plus a variance floor of $0.1$. One sample is taken
per step, $d_y=1$, $d_\xi=1$.

\textbf{Prior.} The three rates are jointly log-normal,
$\log\theta\sim\mathcal N\big((\log 1.0,\log 0.1,\log 20.0),\,0.05\,I_3\big)$, so each
log-parameter is independent normal with variance $0.05$. This matches the
iDAD~\citep{ivanova2021implicit} and Klein~\citep{klein2026supercharging} setups.

\section{Pharmacokinetics: eight-compartment PBPK}
A whole-body, permeability-limited physiologically based pharmacokinetic (PBPK) model: a
twenty-parameter scale-up of the one-compartment task with the same design paradigm
(choose sampling times, observe plasma concentration) but a structured, high-dimensional
parameter space. The body is central blood, seven lumped tissues (liver, kidney, gut,
muscle, adipose, richly-perfused, slowly-perfused), and an oral-dose depot.

\textbf{States and physiology.} Each tissue $i$ is split into a vascular sub-compartment
(concentration $C_{\mathrm{vas},i}$) and a tissue sub-compartment ($C_{\mathrm{tis},i}$),
giving $16$ internal states
$(M_{\mathrm{depot}},C_{\mathrm{blood}},\{C_{\mathrm{vas},i},C_{\mathrm{tis},i}\}_{i=1}^7)$.
Fixed $70$-kg physiology sets the blood volume $V_{\mathrm{blood}}=5.6$\,L, the tissue
volumes $V^{\mathrm{tot}}_i$, the vascular fractions $f_{v,i}$ (so that
$V_{\mathrm{vas},i}=f_{v,i}V^{\mathrm{tot}}_i$ and
$V_{\mathrm{tis},i}=(1-f_{v,i})V^{\mathrm{tot}}_i$), and the tissue blood flows $Q_i$
summing to a $390$\,L/h cardiac output.

\textbf{Forward model.} The inferred parameters are partition coefficients $K_{p,i}$,
permeability-surface products $\mathrm{PS}_i$, hepatic, renal, and gut clearances
$\mathrm{CL}_h,\mathrm{CL}_r,\mathrm{CL}_{\mathrm{gut}}$, oral absorption rate $k_a$,
bioavailability $F\in(0,1)$, and unbound plasma fraction $f_{up}\in(0,1)$. The dynamics
are linear,
\begin{align*}
\dot M_{\mathrm{depot}} &= -k_a M_{\mathrm{depot}},\\
\dot C_{\mathrm{blood}} &= \frac{1}{V_{\mathrm{blood}}}\Big(\textstyle\sum_{i=1}^7 Q_i\big(C_{\mathrm{vas},i}-C_{\mathrm{blood}}\big)+k_a F M_{\mathrm{depot}}-\mathrm{CL}_r f_{up}C_{\mathrm{blood}}\Big),\\
\dot C_{\mathrm{vas},i} &= \frac{1}{V_{\mathrm{vas},i}}\Big(Q_i\big(C_{\mathrm{blood}}-C_{\mathrm{vas},i}\big)-\mathrm{PS}_i\big(f_{up}C_{\mathrm{vas},i}-C_{\mathrm{tis},i}/K_{p,i}\big)\Big),\\
\dot C_{\mathrm{tis},i} &= \frac{1}{V_{\mathrm{tis},i}}\Big(\mathrm{PS}_i\big(f_{up}C_{\mathrm{vas},i}-C_{\mathrm{tis},i}/K_{p,i}\big)-\big(\delta_{i,\mathrm{liver}}\mathrm{CL}_h+\delta_{i,\mathrm{gut}}\mathrm{CL}_{\mathrm{gut}}\big)C_{\mathrm{tis},i}/K_{p,i}\Big),
\end{align*}
where $\delta_{i,\mathrm{liver}}$ and $\delta_{i,\mathrm{gut}}$ are indicators that switch
on hepatic and gut elimination for the liver and gut compartments only. Because the depot
is part of the state, the system is autonomous and linear, $\dot y=A(\theta)y$, and is
solved exactly by the matrix exponential $y(\xi)=\exp\!\big(A(\theta)\,\xi\big)y_0$.
Dosing is fixed: a $100$\,mg intravenous bolus enters blood at $t=0$
($C_{\mathrm{blood}}(0)=100/V_{\mathrm{blood}}$) and a $100$\,mg oral dose loads the depot
($M_{\mathrm{depot}}(0)=100$); the dual route makes $F$ and $k_a$ jointly identifiable
from a single plasma trajectory.

\textbf{Prior.} The $d_\theta=20$ parameters are seven partition coefficients
$K_{p,i}\sim\mathrm{LogNormal}(\log 2,0.3^2)$, seven permeabilities
$\mathrm{PS}_i\sim\mathrm{LogNormal}(\log 50,0.5^2)$, the clearances
$\mathrm{CL}_h\sim\mathrm{LogNormal}(\log 20,0.4^2)$,
$\mathrm{CL}_r\sim\mathrm{LogNormal}(\log 10,0.5^2)$,
$\mathrm{CL}_{\mathrm{gut}}\sim\mathrm{LogNormal}(\log 5,0.5^2)$, the absorption rate
$k_a\sim\mathrm{LogNormal}(\log 1.5,0.4^2)$, and the two fractions $F=\mathrm{sigmoid}(z_F)$
and $f_{up}=\mathrm{sigmoid}(z_{up})$ with $z_F,z_{up}\sim\mathcal N(0,1)$. These tight,
calibrated priors are what give the system a small escape threshold despite $d_\theta=20$.

\textbf{Observation and design.} A single plasma reading $C_{\mathrm{blood}}$ at each
sampling time $\xi\in[0,24]$\,h, $d_y=1$, $d_\xi=1$, with the same heteroscedastic noise
as the one-compartment model, $s=\sqrt{(0.1\,C_{\mathrm{blood}})^2+0.1}$.

\section{PI3K signalling}
A four-state model of iRap-induced PIP3 signalling at the plasma
membrane~\citep{liang2024structural}, read out by FRET microscopy. A chemical dimeriser
(iRap) recruits PI3K to the membrane, which produces PIP3, reported by a translocating
fluorescent probe; an inhibitor (LY29) blocks PI3K. The two designs are the iRap and LY29
doses.

\textbf{States and conservation.} Write the four independent states as
$x=(x_1,x_2,x_3,x_4)$: $x_1$ is the iRap-p85 complex, $x_2$ the membrane-recruited
Lyn-iRap-p85 complex, $x_3$ the second messenger PIP3, and $x_4$ the membrane-bound
YFP-PH probe. Three conservation laws (all totals equal to one, in arbitrary fluorescence
units) give the dependent quantities $c_{\mathrm{cfp}}=1-x_1-x_2$ (free CFP-FKBP-p85),
$\ell=1-x_2$ (free Lyn-FRB), and $x_{\mathrm{cyt}}=1-x_4$ (cytosolic probe).

\textbf{Forward model.} With parameters $\theta=(k_r,k_t,k_{ly},\phi,k_a,\alpha)$ and
designs $u=(u_{\mathrm{iRap}},u_{\mathrm{LY29}})$, the reaction fluxes are
\[
v_{\mathrm{dimer}}=k_r\,c_{\mathrm{cfp}}\,u_{\mathrm{iRap}},\quad
v_{\mathrm{trimer}}=k_t\,\ell\,x_1,\quad
v_{\mathrm{pi3k}}=\frac{x_2}{1+u_{\mathrm{LY29}}/k_{ly}},\quad
v_{\mathrm{deg}}=\phi\,x_3,\quad
v_{\mathrm{ph}}=k_a\,x_{\mathrm{cyt}}\,x_3-\alpha\,x_4,
\]
where $k_r$ is the dimerisation rate, $k_t$ the membrane-recruitment rate, $k_{ly}$ the
LY29 inhibition constant, $\phi$ the PIP3 degradation rate, and $k_a,\alpha$ the probe
binding and unbinding rates. The state derivatives are
\[
\dot x_1=v_{\mathrm{dimer}}-v_{\mathrm{trimer}},\quad
\dot x_2=v_{\mathrm{trimer}},\quad
\dot x_3=v_{\mathrm{pi3k}}-v_{\mathrm{deg}},\quad
\dot x_4=v_{\mathrm{ph}},
\]
from $x(0)=0$.

\textbf{Prior.} The six positive rates $\theta=(k_r,k_t,k_{ly},\phi,k_a,\alpha)$ are
log-normal about the published best-fit values,
\[
\log\theta\sim\mathcal N\big(\log(1.56,0.03,0.16,1.80,0.58,0.11),\,0.5^2 I_6\big).
\]

\textbf{Observation and design.} The two measured channels are the cytosolic fluorescence
readouts $(c_{\mathrm{cfp}},x_{\mathrm{cyt}})$, $d_y=2$, with additive noise
$\mathcal N(0,0.02^2)$; the designs are the iRap dose $u_{\mathrm{iRap}}\in[0,1]\,\mu$M
and the LY29 dose $u_{\mathrm{LY29}}\in[0,25]\,\mu$M ($d_\xi=2$), applied as eight
piecewise-constant three-minute intervals over $24$ minutes.

\section{Monod bioreactor}
A fed-batch bioreactor with Monod growth kinetics~\citep{monod1949growth,strouwen2026deep}:
a microbial culture consumes a feed substrate, and the single design is the feed rate. The
dynamics are a mildly stiff ODE.

\textbf{Forward model.} The state $(C_s,C_x,V)$ is the substrate concentration (g/L), the
biomass concentration (g/L), and the broth volume (L). The specific growth rate is the
Monod form $\mu=\mu_{\max}C_s/(K_s+C_s)$, and with fixed yield $Y_{xs}=0.777$, fixed feed
concentration $C_{s,\mathrm{in}}=50$\,g/L, and feed rate $\xi=Q_{\mathrm{in}}\in[0,6]$\,L/h,
\[
\dot C_s=-\frac{\mu}{Y_{xs}}C_x+\frac{Q_{\mathrm{in}}}{V}\big(C_{s,\mathrm{in}}-C_s\big),
\qquad
\dot C_x=\mu C_x-\frac{Q_{\mathrm{in}}}{V}C_x,
\qquad
\dot V=Q_{\mathrm{in}},
\]
from $(C_s,C_x,V)(0)=(3,0.3,7)$, run as $14$ hourly steps.

\textbf{Prior.} $d_\theta=2$, uniform: $\mu_{\max}\sim\mathrm{Uniform}(0.3,0.5)$ and
$K_s\sim\mathrm{Uniform}(0.3,0.6)$.

\textbf{Observation and design.} $d_y=1$ (the substrate $C_s$) with $\mathcal N(0,0.1^2)$
noise; $d_\xi=1$. The weak sensitivity of $C_s$ to $(\mu_{\max},K_s)$ over feasible feed
rates gives Monod a low information content, which is why it is a predicted counter-case
where a trajectory policy can still help.

\section{Goodwin oscillator}
A three-variable gene-regulatory oscillator~\citep{strouwen2026deep}: an mRNA drives a
protein, which drives a downstream species that represses transcription through a steep
Hill term. The design is an external transcriptional input.

\textbf{Forward model.} The state $(X,Y,Z)$ is the mRNA, protein, and downstream product.
With parameters $\theta=(V_s,K,n,k_d,k_t)$ and scalar input $u\in[0,1]$,
\[
\dot X=\frac{V_s\,K^n}{K^n+Z^n}-k_d X+u,
\qquad
\dot Y=k_t X-k_d Y,
\qquad
\dot Z=k_t Y-k_d Z,
\]
from $(X,Y,Z)(0)=(0.1,0.1,0.1)$. Here $V_s$ is the maximum transcription rate, $K$ the
repression threshold, $n$ the Hill cooperativity, $k_d$ a shared degradation rate, and
$k_t$ the translation and conversion rate.

\textbf{Prior.} The parameters are stored in the order $(V_s,K,n,k_d,k_t)$, $d_\theta=5$.
The four positive rates are log-normal about $(V_s,K,k_d,k_t)=(0.5,1.0,0.1,1.0)$ with
log-scale $0.4$, that is $\log V_s\sim\mathcal N(\log 0.5,0.4^2)$ and likewise for
$K,k_d,k_t$. The Hill exponent is $n=\max(12+1.5\,z,\,7)$ with $z\sim\mathcal N(0,1)$: a
normal centred at $12$ with scale $1.5$, floored at $7$ to keep the system in its
oscillatory regime. The high cooperativity produces a steep, nearly flat identifiability
landscape, which makes Goodwin a predicted counter-case.

\textbf{Observation and design.} $d_y=1$ (the mRNA $X$) with $\mathcal N(0,0.05^2)$ noise;
$d_\xi=1$.

\section{ASM1 wastewater}
Activated Sludge Model No.~1~\citep{henze2000activated}, the standard biokinetic model of
a wastewater aeration tank. We use a single continuous-flow stirred-tank configuration in
which $13$ state concentrations evolve under eight microbial processes, with the aeration
rate and the sludge wastage as designs. The ODE is stiff and is integrated with an
implicit solver.

\textbf{States.} The state vector is
\[
(S_I,S_S,X_I,X_S,X_{BH},X_{BA},X_P,S_O,S_{NO},S_{NH},S_{ND},X_{ND},S_{ALK}),
\]
comprising inert and readily-biodegradable soluble COD ($S_I,S_S$); inert,
slowly-biodegradable, heterotroph, autotroph, and inert-particulate COD
($X_I,X_S,X_{BH},X_{BA},X_P$); dissolved oxygen ($S_O$); nitrate and ammonia nitrogen
($S_{NO},S_{NH}$); soluble and particulate organic nitrogen ($S_{ND},X_{ND}$); and
alkalinity ($S_{ALK}$). All concentrations are in mg/L.

\textbf{Process rates.} With Monod saturation switches, the eight process rates are
\begin{align*}
\rho_0 &= \mu_H\tfrac{S_S}{K_S+S_S}\tfrac{S_O}{K_{OH}+S_O}X_{BH}, &
\rho_1 &= \mu_H\tfrac{S_S}{K_S+S_S}\tfrac{K_{OH}}{K_{OH}+S_O}\tfrac{S_{NO}}{K_{NO}+S_{NO}}\eta_g X_{BH},\\
\rho_2 &= \mu_A\tfrac{S_{NH}}{K_{NH}+S_{NH}}\tfrac{S_O}{K_{OA}+S_O}X_{BA}, &
\rho_3 &= b_H X_{BH},\quad \rho_4=b_A X_{BA},\quad \rho_5=k_a S_{ND}X_{BH},\\
\rho_6 &= k_h\tfrac{X_S/X_{BH}}{K_X+X_S/X_{BH}}\big(\tfrac{S_O}{K_{OH}+S_O}+\eta_h\tfrac{K_{OH}}{K_{OH}+S_O}\tfrac{S_{NO}}{K_{NO}+S_{NO}}\big)X_{BH}, &
\rho_7 &= \rho_6\,X_{ND}/X_S,
\end{align*}
the aerobic and anoxic heterotrophic growth, autotrophic growth, heterotrophic and
autotrophic decay, ammonification, and the two hydrolysis processes.

\textbf{Mass balances.} The processes enter the $13$ states through the ASM1 Petersen
stoichiometry, together with a continuous-flow exchange at $Q/V=4\,\mathrm{d}^{-1}$ (a
fixed $6$\,h hydraulic residence time), aeration $k_La(S_{O,\mathrm{sat}}-S_O)$ added to
oxygen with saturation $S_{O,\mathrm{sat}}=8$, and a wastage term $Q_w$ on the
particulate states:
\begin{align*}
\dot S_I &= \tfrac{Q}{V}\big(S_I^{\mathrm{in}}-S_I\big),\\
\dot S_S &= -\tfrac{1}{Y_H}\rho_0-\tfrac{1}{Y_H}\rho_1+\rho_6+\tfrac{Q}{V}\big(S_S^{\mathrm{in}}-S_S\big),\\
\dot X_I &= \tfrac{Q}{V}X_I^{\mathrm{in}}-Q_w X_I,\\
\dot X_S &= (1-f_P)\rho_3+(1-f_P)\rho_4-\rho_6+\tfrac{Q}{V}X_S^{\mathrm{in}}-Q_w X_S,\\
\dot X_{BH} &= \rho_0+\rho_1-\rho_3-Q_w X_{BH},\qquad
\dot X_{BA} = \rho_2-\rho_4-Q_w X_{BA},\qquad
\dot X_P = f_P\rho_3+f_P\rho_4-Q_w X_P,\\
\dot S_O &= -\tfrac{1-Y_H}{Y_H}\rho_0-\tfrac{4.57-Y_A}{Y_A}\rho_2+k_La\big(S_{O,\mathrm{sat}}-S_O\big)-\tfrac{Q}{V}S_O,\\
\dot S_{NO} &= -\tfrac{1-Y_H}{2.86\,Y_H}\rho_1+\tfrac{1}{Y_A}\rho_2-\tfrac{Q}{V}S_{NO},\\
\dot S_{NH} &= -i_{XB}\rho_0-i_{XB}\rho_1-\big(i_{XB}+\tfrac{1}{Y_A}\big)\rho_2+\rho_5+\tfrac{Q}{V}\big(S_{NH}^{\mathrm{in}}-S_{NH}\big),\\
\dot S_{ND} &= -\rho_5+\rho_7+\tfrac{Q}{V}\big(S_{ND}^{\mathrm{in}}-S_{ND}\big),\\
\dot X_{ND} &= (i_{XB}-f_P i_{XP})\rho_3+(i_{XB}-f_P i_{XP})\rho_4-\rho_7+\tfrac{Q}{V}X_{ND}^{\mathrm{in}}-Q_w X_{ND},\\
\dot S_{ALK} &= -\tfrac{i_{XB}}{14}\rho_0+\Big(\tfrac{1-Y_H}{14\cdot2.86\,Y_H}-\tfrac{i_{XB}}{14}\Big)\rho_1-\big(\tfrac{i_{XB}}{14}+\tfrac{1}{7\,Y_A}\big)\rho_2+\tfrac{1}{14}\rho_5+\tfrac{Q}{V}\big(S_{ALK}^{\mathrm{in}}-S_{ALK}\big).
\end{align*}

\textbf{Fixed coefficients.} $Y_A=0.24$, $f_P=0.08$, $i_{XB}=0.086$, $i_{XP}=0.06$,
$K_{NO}=0.5$, $K_{OA}=0.4$, $\eta_h=0.4$, and $S_{O,\mathrm{sat}}=8$. The fixed influent
concentrations are $(S_I,S_S,X_I,X_S,S_{NH},S_{ND},X_{ND},S_{ALK})^{\mathrm{in}}=
(30,70,50,200,30,7,10,7)$, with influent oxygen and nitrate zero, and the initial state
is $x_0=(30,5,50,80,2000,150,100,2,8,2,1,5,7)$.

\textbf{Prior.} The $d_\theta=12$ kinetic parameters are independently log-normal about
their nominal values, $\log\theta_j\sim\mathcal N(\log m_j,\sigma_j^2)$. The nominal value
$m_j$ and log-scale $\sigma_j$ of each, written as $\theta_j:(m_j,\sigma_j)$, are
\[
\begin{array}{llll}
\mu_H:(6.0,\,0.45) & K_S:(20,\,0.62) & K_{OH}:(0.2,\,0.58) & \mu_A:(0.8,\,0.40)\\[2pt]
K_{NH}:(1.0,\,0.58) & b_H:(0.62,\,0.45) & b_A:(0.15,\,0.45) & \eta_g:(0.8,\,0.23)\\[2pt]
k_h:(3.0,\,0.52) & K_X:(0.03,\,0.58) & k_a:(0.08,\,0.58) & Y_H:(0.67,\,0.19).
\end{array}
\]

\textbf{Observation and design.} $d_y=4$, the effluent quality variables
$(S_S,S_{NH},S_{NO},S_O)$ with heteroscedastic noise $s=\sqrt{(0.05\,y)^2+0.5^2}$ (five
percent proportional plus a $0.5$ floor); the designs are the aeration coefficient
$k_La\in[10,300]\,\mathrm{d}^{-1}$ and the waste-sludge fraction
$Q_w\in[0.01,0.2]\,\mathrm{d}^{-1}$, $d_\xi=2$. The stiffness and the resulting adjoint
cost are what make policy training infeasible here.

\section{Multi-region SEIR}
An epidemic model over $R$ coupled regions, each an SEIR compartment model, with
per-region interventions as designs. It is the high-information, stiff dynamical anchor of
the suite.

\textbf{Forward model.} Region $r$ has states $(S_r,E_r,I_r,R_r)$, the susceptible,
exposed, infectious, and removed counts, with a fixed population $N=10^4$. Infection
couples across regions through $\mathrm{eff}_r=I_r+\kappa\sum_{j\ne r}I_j$, a weighted sum
of local and imported infectious counts with coupling strength $\kappa$. With shared
incubation rate $\sigma$, recovery rate $\gamma$, per-region transmission rates $\beta_r$,
and intervention efficacies $\varepsilon_v,\varepsilon_d,\varepsilon_t$ (vaccination,
distancing, treatment), and per-region designs $(v_r,d_r,\tau_r)$,
\begin{align*}
\dot S_r &= -\beta_r(1-\varepsilon_d d_r)\,S_r\,\mathrm{eff}_r/N-v_r\varepsilon_v S_r, &
\dot E_r &= \beta_r(1-\varepsilon_d d_r)\,S_r\,\mathrm{eff}_r/N-\sigma E_r,\\
\dot I_r &= \sigma E_r-\gamma I_r-\tau_r\varepsilon_t I_r, &
\dot R_r &= \gamma I_r+\tau_r\varepsilon_t I_r+v_r\varepsilon_v S_r,
\end{align*}
from $(S_r,E_r,I_r,R_r)(0)=(N-I_0,0,I_0,0)$ with $I_0=10$, integrated in steps of two
days.

\textbf{Prior.} $d_\theta=6+R$. The shared rates are
\[
\sigma\sim\mathrm{LogNormal}(\log 0.2,0.3^2),\quad
\gamma\sim\mathrm{LogNormal}(\log 0.1,0.3^2),\quad
\kappa\sim\mathrm{LogNormal}(\log 0.05,0.5^2),
\]
the per-region transmission is $\beta_r\sim\mathrm{LogNormal}(\log 0.4,0.5^2)$, and the
efficacies are $\varepsilon_v,\varepsilon_d,\varepsilon_t\sim\mathrm{Beta}(2,2)$.

\textbf{Observation and design.} The observation is the exposed and infectious counts of
every region, stacked as $(E_1,\dots,E_R,I_1,\dots,I_R)$, $d_y=2R$, measured with
multiplicative lognormal noise $y=\mu\,e^{0.15\,\epsilon}$, $\epsilon\sim\mathcal N(0,1)$,
where $\mu$ is the clean count. The design is $(v_r,d_r,\tau_r)_{r=1}^R$, $d_\xi=3R$, with
$v_r\in[0,0.05]$, $d_r\in[0,1]$, $\tau_r\in[0,0.5]$. We report $R=3$ as the main system
($d_\theta=9$, $d_\xi=9$, $d_y=6$); the larger $R=8$ appears only as a high-dimensional
diagnostic in the saturation survey ($d_\theta=14$, $d_\xi=24$, $d_y=16$;
Table~\ref{tab:escape-full}).

\section{Synthetic Michaelis-Menten cascade (controlled toy)}
This is the controlled system behind the cost-shift demonstration of
Section~\ref{sec:r-toy} and Figure~\ref{fig:toy}. It is a synthetic three-species enzyme
cascade, chosen to be small enough to admit a near-exact reference posterior by prior
importance sampling, yet to reproduce the saturation regime in kind: kinetic rate
constants, a four-order-of-magnitude prior, lognormal noise, and a multi-step trajectory.

\textbf{Forward model.} Three species react in sequence $A\to B\to C$ through saturating
Michaelis-Menten steps. With maximal rates $V_1,V_2,V_3$, half-saturation constants
$K_1,K_2,K_3$, and fluxes $f_1=V_1A/(K_1+A)$, $f_2=V_2B/(K_2+B)$, $f_3=V_3C/(K_3+C)$,
\[
\dot A=-f_1,\qquad \dot B=f_1-f_2,\qquad \dot C=f_2-f_3,
\]
from initial condition $A(0)=\xi$, $B(0)=C(0)=0$. The single design $\xi=A(0)\in[0.1,100]$
sets the initial substrate amount ($d_\xi=1$). The three species are observed at $T=10$
evenly spaced times in $(0,5]$ with multiplicative lognormal noise
$y=\mu\,e^{0.1\,\epsilon}$, $\epsilon\sim\mathcal N(0,1)$, so $d_y=3$ channels per time.

\textbf{Prior.} The six parameters are inferred in log space, $\theta=(\log V_1,\log K_1,
\log V_2,\log K_2,\log V_3,\log K_3)$, each independent and uniform over four orders of
magnitude, $\log\theta_j\sim\mathrm{Uniform}(\log 0.01,\log 100)$; hence $d_\theta=6$.

\textbf{Configuration.} The cost-shift demonstration uses a prior-contrastive budget
$L=2000$ and a posterior-contrastive budget $J=200$, sweeping the NPE training budget
$N_{\mathrm{NPE}}$ from $10^2$ to $5\times10^4$ (Figure~\ref{fig:toy}). The toy is not part
of the deployed benchmark suite of Table~\ref{tab:config}.

\begin{table}[h]
\centering\footnotesize
\caption[Experiment configurations]{Per-system experiment configurations for the
deployed PACE-ftk runs. Optimiser is forward-only cross-entropy (CEM); $L/J$ is the
contrastive budget ($J$ particles for PACE, $L_{\mathrm{train}}$ for DAD);
$N_{\mathrm{cand}}$ is the candidate count per CEM round; $C$ is the number of CEM
refinement rounds ($C{=}0$ is an unrefined random pool). The PACE-grad ablation uses the
same proposal and candidate budget with gradient refinement instead of CEM.}
\label{tab:config}
\begin{tabular}{lrrrlrrrr}
\toprule
System & $d_\theta$ & $d_\xi$ & $T$ & opt. & $L/J$ & $N_{\mathrm{cand}}$ & $C$ & $L_{\mathrm{eval}}$ \\
\midrule
CES & 4 & 6 & 10 & CEM & 1000 & 1000 & 1 & $10^7$ \\
ASM1 & 12 & 2 & 5 & CEM & 500 & 50 & 0 & $5{\times}10^4$ \\
SEIR R=3 & 9 & 9 & 10 & CEM & 1000 & 1000 & 1 & $10^7$ \\
Monod & 2 & 1 & 14 & CEM & 1000 & 1000 & 1 & $5{\times}10^3$ \\
Goodwin & 5 & 1 & 10 & CEM & 1000 & 1000 & 1 & $5{\times}10^3$ \\
Src.\ loc.\ $p{=}2$ & 4 & 2 & 10 & CEM & 1000 & 1000 & 1 & $10^5$ \\
Src.\ loc.\ $p{=}5$ & 10 & 5 & 30 & CEM & 1000 & 1000 & 1 & $10^5$ \\
Src.\ loc.\ $p{=}8$ & 16 & 8 & 30 & CEM & 1000 & 1000 & 1 & $10^5$ \\
PK 1-comp $T{=}3$ & 3 & 1 & 3 & CEM & 1000 & 1000 & 1 & $2{\times}10^3$ \\
PK 1-comp $T{=}5$ & 3 & 1 & 5 & CEM & 1000 & 1000 & 1 & $2{\times}10^3$ \\
PK 8-comp & 20 & 1 & 5 & CEM & 1000 & 1000 & 1 & $2{\times}10^3$ \\
PI3K & 6 & 2 & 8 & CEM & 1000 & 1000 & 1 & $10^5$ \\
\bottomrule
\end{tabular}
\end{table}

\section{Baseline policy (DAD) configuration}
The DAD baseline follows \citet{foster2021deep}: a permutation-invariant history encoder
(a shared per-step MLP applied to each $(\xi_i,y_i)$ pair, sum-pooled across steps)
feeding a design-emitter MLP that outputs the next design. It is trained by Adam
stochastic gradient ascent on the trajectory sPCE bound (Eq.~\eqref{eq:traj-spce}) with
$L_{\mathrm{train}}$ prior contrastives, $L_{\mathrm{train}}=1000$ for all systems except
CES ($1024$), with per-system gradient-step budgets of $3{,}000$ to $50{,}000$ and batch
sizes of $8$ to $128$; the resulting training times are in Table~\ref{tab:walltime}. For
SEIR R=3, $L_{\mathrm{train}}$ is additionally swept over
$\{10^2,10^3,10^4,5{\times}10^4,10^5\}$ (Table~\ref{tab:dad-sweep}). DAD is evaluated on
the same $\theta_{\mathrm{true}}$ seeds and at the same $L_{\mathrm{eval}}$ as PACE. On
CES, training diverges to NaN before convergence (the censored-likelihood instability
also reported by \citet{hedman2025stepdad}), and the reported result uses the best
checkpoint before divergence; on ASM1, and on SEIR R=3 at the budget needed to escape
saturation, training is infeasible (stiff-ODE adjoint with
$L_{\mathrm{escape},50}$ beyond any feasible $L_{\mathrm{train}}$;
Section~\ref{sec:m-forward}).

\chapter{Neural posterior estimator}
\label{app:npe}

\section{Data generation and posterior-target correctness}
Training data consist of $N$ triples $(\theta_i,\xi^{(i)}_{1:T},y^{(i)}_{1:T})$:
$\theta_i\sim p(\theta)$, designs $\xi^{(i)}_t\sim\mathrm{Uniform}(\Xi)$ sampled
independently for all $T$ steps, and observations
$y^{(i)}_t\sim p(y_t\mid\theta_i,\xi^{(i)}_t)$. Because the design distribution
$p(\xi_{1:t})$ does not depend on $\theta$, the training-joint posterior equals the
true posterior:
\begin{equation}
\label{eq:npe-correct}
p_{\mathrm{train}}(\theta\mid h_t)
= \frac{p(\theta)\,p(\xi_{1:t})\,p(y_{1:t}\mid\theta,\xi_{1:t})}{p_{\mathrm{train}}(h_t)}
= \frac{p(\theta)\,p(y_{1:t}\mid\theta,\xi_{1:t})}{p(y_{1:t}\mid\xi_{1:t})}
= p(\theta\mid y_{1:t},\xi_{1:t}),
\end{equation}
where $p(\xi_{1:t})$ cancels. The NPE therefore targets the correct posterior for
\emph{any} history $h_t$, regardless of how the designs were chosen. At deployment PACE
queries the NPE with histories containing optimally selected designs; the posterior is
determined by Bayes' rule given the data, not by the training design distribution. Each
training sample is a full $T$-step trajectory, and random truncation
($t_i\sim\mathrm{Uniform}\{1,\dots,T\}$, masking steps $t>t_i$) teaches the network to
produce posteriors from any prefix length.

\section{Set-equivariant conditioning}
The posterior $p(\theta\mid h_t)$ is permutation-invariant with respect to the elements
of $h_t=\{(y_s,\xi_s,s)\}_{s=1}^{t}$, and the conditioning architecture reflects this.
Each element is encoded as a triple $(y_s,\xi_s,s/T)$: a $z$-scored observation
(log-transformed under lognormal noise), a $z$-scored design, and a normalised time
index. A shared trial network $f_{\mathrm{trial}}:\mathbb{R}^{d_y+d_\xi+1}\to\mathbb{R}^{d_e}$
(2 layers, SiLU, $d_e=64$) processes each triple independently,
$e_s=f_{\mathrm{trial}}(y_s,\xi_s,s/T)$. The embeddings are sum-pooled,
$\bar e=\sum_{s=1}^{t}e_s$ (sum, not mean, to preserve the magnitude signal so that $t$
observations produce a different embedding than one), a count feature $t/T$ is appended,
and a set network $f_{\mathrm{set}}:\mathbb{R}^{d_e+1}\to\mathbb{R}^{d_{\mathrm{emb}}}$
(2 layers, SiLU, $d_{\mathrm{emb}}=128$) produces the final conditioning vector.

\section{Density estimator and training}
$q_\phi(\theta\mid h_t)$ is a neural spline flow~\citep{durkan2019neural} with $12$
coupling transforms and hidden dimension $256$. Parameters are mapped to an
unconstrained space (log for $(0,\infty)$, logit for $(0,1)$). NSF is chosen over a
flow-matching estimator because the importance reweighting requires fast, exact
evaluation of $\log q_\phi$ ($\sim$1\,ms, versus $\sim$500\,ms for the ODE solve of a
flow-matching posterior). Training uses the \texttt{sbi} library~\citep{tejerocantero2020sbi}
with \texttt{z\_score\_x=`none'}, because the flattened condition includes binary mask
channels that $z$-scoring would corrupt. The hyperparameters are:
\begin{center}\small
\begin{tabular}{ll}
\toprule
Hyperparameter & Value \\
\midrule
Coupling transforms / hidden dim & 12 / 256 \\
Learning rate / batch size & $5\times10^{-4}$ / 256 \\
Max epochs / early-stopping patience & 500 / 30 epochs \\
Trial network & 2 layers, 64 hidden, SiLU \\
Set network & 2 layers, 128 hidden, SiLU \\
\bottomrule
\end{tabular}
\end{center}
